---
title: "Lab 01: COVID Data"
subtitle: "ESS 523c"
author:
  name: Kevin
  email: kevin.horrell@colostate.edu
format: html
---


```{r, echo = F, message = F, warning = F}
library(tidyverse)
library(flextable)
library(zoo)
```


Question 1: Daily Summary
To start, you should set up a reproducible framework to communicate the following in a way that can be updated 
every time new data is released (daily):
    1. cumulative cases in the 5 worst counties
    2. total NEW cases in the 5 worst counties
    3. A list of safe counties

You should build this analysis so that running it will extract the most current data straight from the NY-Times URL and the state name and date are parameters that can be changed allowing this report to be run for other states/dates.

Read in Data
```{r, echo = T}
dat <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
```

Initial Objects
```{r}
my.date <- as.Date("2022-02-01")
my.state <- "Colorado"
```

Colorado Data
```{r}
co_dat <- dat %>%
  filter(state == my.state) %>%
  group_by(county) %>%
  mutate(new_cases = cases - lag(cases, n = 1),
         new_deaths = deaths - lag(deaths, n = 1)) %>%
  drop_na() %>%
  ungroup()
```

Data Tables
```{r}
co_dat_cum <- co_dat %>%
  filter(date == my.date)

slice_max(co_dat_cum, n = 5, order_by = new_cases) %>%
  select(county, state, cases, new_cases) %>%
  flextable() %>%
  set_caption("Top 5 Counties with New Cases")

slice_max(co_dat_cum, n = 5, order_by = new_deaths) %>%
  select(county, state, cases, deaths, new_deaths) %>%
  flextable() %>%
  set_caption("Top 5 Counties with Most Deaths") 
  
```

Question 2: Evaluating Census Data
